---
title: "Population Change in the Palatinate"
author: "Anna Duurland"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 5,
  fig.align = "center")

library(tidyverse)
library(sf)
library(plotly)
library(DT)
```

```{r data}
df <- read.csv("population_palatinate.csv")

geo <- st_read("palatinate_municipalities.geojson", quiet = TRUE)
```

# Exploring Population Trends

Which municipalities experienced the strongest population growth or decline between 1970 and 2022?

```{r population-change}
population_change <- df %>%
  filter(
    !is.na(year),
    !is.na(total_population),
    !is.na(gmdesch)
  ) %>%
  arrange(municipality, year) %>%
  group_by(municipality) %>%
  summarise(
    start_year = first(year),
    end_year = last(year),
    start_population = first(total_population),
    end_population = last(total_population),
    change_percent = (end_population - start_population) /
      start_population * 100,
    .groups = "drop"
  )

winners <- population_change %>%
  slice_max(change_percent, n = 3)

losers <- population_change %>%
  slice_min(change_percent, n = 3)

winners

losers
```

# Exploring Long-Term Population Decline

What would happen if historical population trends continued linearly?

```{r population-projection}
population_projection <- df %>%
  filter(
    !is.na(year),
    !is.na(total_population),
    !is.na(gmdesch)
  ) %>%
  group_by(municipality) %>%
  filter(n() >= 2) %>%
  summarise(
    intercept = coef(lm(total_population ~ year))[1],
    slope = coef(lm(total_population ~ year))[2],
    .groups = "drop"
  ) %>%
  filter(slope < 0) %>%
  mutate(
    projected_zero_year = -intercept / slope
  ) %>%
  filter(
    projected_zero_year > max(df$year, na.rm = TRUE)
  ) %>%
  arrange(projected_zero_year)

earliest_zero <- population_projection %>%
  slice_head(n = 3)

earliest_zero %>%
  mutate(
    projected_zero_year = round(projected_zero_year)
  )
```

These projections are exploratory rather than demographic forecasts. They show when the linear trend would theoretically reach zero if the historical trend continued unchanged.

# Population Trends

## Strongest Growth and Decline

The following visualization compares the three municipalities with the strongest population growth with the three experiencing the strongest decline.

```{r winners-losers-data}
selected_change <- c(
  winners$municipality,
  losers$municipality
)

trend_change <- df %>%
  filter(
    municipality %in% selected_change,
    !is.na(year),
    !is.na(total_population)
  ) %>%
  arrange(municipality, year) %>%
  group_by(municipality) %>%
  mutate(
    population_index = total_population /
      first(total_population) * 100
  ) %>%
  ungroup()
```

```{r interactive-bar}

top_population_change <- bind_rows(
  population_change %>% slice_min(change_percent, n = 10),
  population_change %>% slice_max(change_percent, n = 10)
)

p_bar <- ggplot(
  top_population_change,
  aes(
    x = reorder(municipality, change_percent),
    y = change_percent,
    fill = change_percent > 0,
    text = paste0(
      municipality, ": ",
      round(change_percent, 1), "%"
    )
  )
) +
  geom_col(width = 0.7) +
  coord_flip() +
  geom_hline(yintercept = 0) +
  labs(
    title = "Growing and Shrinking Municipalities",
    subtitle = "Municipalities with the strongest population change, 1970–2022",
    x = NULL,
    y = "Population change (%)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none"
  )

ggplotly(
  p_bar,
  tooltip = "text"
)
```


```{r interactive-winners-losers}
p_change <- ggplot(
  trend_change,
  aes(
    x = year,
    y = population_index,
    color = municipality,
    group = municipality,
    text = paste0(
      municipality,
      "<br>Year: ", year,
      "<br>Index: ", round(population_index, 0)))) +
  geom_line(linewidth = 1.2) +
  geom_hline(
    yintercept = 100,
    color = "grey60",
    linetype = "dotted") +
  labs(
    title = "Growing and Shrinking Municipalities",
    subtitle = "Population development since 1970",
    x = NULL,
    y = "Index (1970 = 100)",
    color = NULL) +
  theme_minimal() +
  theme(
    legend.position = "right")

ggplotly(
  p_change,
  tooltip = "text")
```

## Municipalities Approaching Zero First

The next visualization shows the three municipalities whose linear population trends reach zero earliest.

```{r projected-zero-data}
selected_projection <- earliest_zero$municipality

trend_projection <- df %>%
  filter(
    municipality %in% selected_projection,
    !is.na(year),
    !is.na(total_population))
```

```{r interactive-projection}
p_projection <- ggplot(
  trend_projection,
  aes(
    x = year,
    y = total_population,
    color = municipality,
    group = municipality,
    text = paste0(
      municipality,
      "<br>Year: ", year,
      "<br>Population: ", total_population))) +
  geom_line(linewidth = 1.2) +
  labs(
    title = "Three Municipalities with the Earliest Projected Zero Population",
    subtitle = "Historical population development",
    x = NULL,
    y = "Population",
    color = NULL) +
  theme_minimal() +
  theme(
    legend.position = "right")

ggplotly(
  p_projection,
  tooltip = "text")
```

# Interactive Table

```{r interactive-table}
population_change %>%
  arrange(change_percent) %>%
  mutate(change_percent = round(change_percent, 1)) %>%
  datatable(
    colnames = c(
      "Municipality",
      "Start Year",
      "End Year",
      "Start Population",
      "End Population",
      "Change (%)"),
    options = list(
      pageLength = 15),
    rownames = FALSE)
```

# Population Across the Palatinate

The map shows population levels across municipalities in the Palatinate at selected points between 1970 and 2022.

```{r interactive-map-slider, fig.width=9, fig.height=7}

slider_years <- c(1970, 1980, 1990,2000, 2010, 2022)

geo_web <- geo %>%
  st_transform(25832) %>%
  st_simplify(
    dTolerance = 250,
    preserveTopology = TRUE) %>%
  st_transform(4326) %>%
  st_make_valid()

geo_web <- suppressWarnings(
  st_collection_extract(
    geo_web,
    "POLYGON"))

geo_web <- st_cast(
  geo_web,
  "MULTIPOLYGON",
  warn = FALSE)

coords <- st_coordinates(geo_web) %>%
  as.data.frame()

coords$gmdesch <- geo_web$gmdesch[coords$L3]

slider_data <- df %>%
  filter(
    year %in% slider_years,
    !is.na(total_population),
    !is.na(gmdesch)) %>%
  select(
    gmdesch,
    municipality,
    year,
    total_population)

plot_data <- coords %>%
  left_join(
    slider_data,
    by = "gmdesch") %>%
  filter(
    !is.na(year),
    !is.na(total_population)) %>%
  mutate(
    hover = paste0(
      "<b>", municipality, "</b>",
      "<br>Year: ", year,
      "<br>Population: ",
      scales::comma(total_population)),
    polygon_id = interaction(
      gmdesch,
      L1,
      L2,
      drop = TRUE))

p_map <- ggplot(
  plot_data,
  aes(x = X, y = Y,
    group = polygon_id,
    frame = year)) +
  geom_polygon(
    aes(
      fill = total_population,
      text = hover),
    color = "white",
    linewidth = 0.05) +
  scale_fill_gradientn(
    colours = c(
      "#8B0000",
      "#D73027",
      "#F46D43",
      "#F7F7F7",
      "#6BAED6",
      "#2171B5",
      "#084594"
    ),
    trans = "log1p",
    na.value = "grey85",
    labels = scales::label_number(
      big.mark = ",")) +
  coord_equal() +
  labs(
    title = "Population Across the Palatinate",
    subtitle = "1970–2022",
    fill = "Population") +
  theme_void()

interactive_map <- ggplotly(
  p_map,
  tooltip = "text") %>%
  animation_opts(
    frame = 500,
    transition = 0,
    redraw = TRUE) %>%
  animation_slider(
    currentvalue = list(
      prefix = "Year: "))

interactive_map
```